---
title: "Periodicity of Yeast Gene Expression"
author: "Yuanhao"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Minimal Simulation Study of the Rank-based Frequency Estimator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette compares the rank-based periodicity test with the likelihood ratio test statistic,
which is derived from the harmonic regression model with Gaussian erros.
Two yeast genes identified as yc1055w and ymr198w are used.
The dataset is available in the `harper` package.

```{r gene, fig.show ='hold',fig.cap = "The observed gene expressions", fig.width=7,fig.height=3}
#----------Read Data----------#
library(harper)
data(alpha)
str(alpha)

y1 <- alpha[row.names(alpha)=="YCL055W",]
y2 <- alpha[row.names(alpha)=="YMR198W",]
n <- length(y1)
t <- 1:n

par(mfrow=c(1,2))
plot(t,y1,type="p",pch=19,
     xlab = "time", ylab = "gene expression",
     main="YCL055W")
plot(t,y1,type="p",pch=19,
     xlab = "time", ylab = "gene expression",
     main="YMR198W")
par(mfrow=c(1,1))
```


## Empirical Distribution
For illustrative purpose, we use the empirical distributions of both test statistics
estimated using only 500 Monte Carlo simualtions.

```{r dist}
n <- 18
M <- 500
lambdaRange <- (ceiling(1001/n):500)/1001
Z <- matrix(rnorm(n*M),ncol=M)
#Takes 168s
set.seed(193)
system.time({
  distLS  <- ptestReg(Z,t=t,method = "LS",
                  lambdalist=lambdaRange,
                  returnPvalue = FALSE)$obsStat
})

system.time({
  distRankLS  <- GetFitRankLS(Z,t=1:n,
                              lambdaRange,
                              phiRange = seq(0,0.99,by = 0.01),
                              InvPI=FALSE,
                              Exact=TRUE)[1,]
})
```


## Harmonic Regression 

We firstly estimate the frequency by the least-square methods on the series,
and then fit the series using the harmonic regression model.

```{r LS}
#----------Fit the harmonic regression----------#
## Estimate the frequency (LS)
lambdaRange <- (ceiling(1001/n):500)/1001
system.time(
LSFit1 <- ptestReg(y1,t=t,method = "LS",
                  lambdalist=lambdaRange,
                  returnPvalue = FALSE)
)
system.time(
LSFit2 <- ptestReg(y2,t=t,method = "LS",
                  lambdalist=lambdaRange,
                  returnPvalue = FALSE)
)
LSFit1$freq #gene YCL055W
LSFit2$freq #gene YMR198W

## Modeling given the estimated frequency
X1 <- data.frame(y1=y1,
                x1=cos(2*pi*LSFit1$freq*t),
                x2=sin(2*pi*LSFit1$freq*t))
X2 <- data.frame(y2=y2,
                x1=cos(2*pi*LSFit2$freq*t),
                x2=sin(2*pi*LSFit2$freq*t))
fitHar1 <- glm(y1~x1+x2,data=X1,family=gaussian())  
fitHar2 <- glm(y2~x1+x2,data=X2,family=gaussian()) 

## Fitted values at continuous time
newt <- seq(t[1],t[length(t)],0.1)
newX1 <- data.frame(x1=cos(2*pi*LSFit1$freq*newt),
                   x2=sin(2*pi*LSFit1$freq*newt))
newX2 <- data.frame(x1=cos(2*pi*LSFit2$freq*newt),
                   x2=sin(2*pi*LSFit2$freq*newt))
predHar1 <- predict(fitHar1,newX1,type="response")
predHar2 <- predict(fitHar2,newX2,type="response")

## P-values of the likelihood ratio test
pvalueLS1 <- mean(distLS>=LSFit1$obsStat) 
pvalueLS2 <- mean(distLS>=LSFit2$obsStat) 
```

The estimated frequencies using the LS method are `r LSFit1$freq` and `r LSFit2$freq`
for the gene YCL055W and the gene YMR198W, respectively.
The corresponding p-values are `r pvalueLS1` and `r pvalueLS2`,respectively.

## Semiparametric Harmonic Regression 

To fit the semiparametric harmonic regression model,
We firstly estimate the frequency by the rank-based methods on the series.
We then fit the series the semiparametric harmonic regression with monotone generalzied additive model.

```{r Rank}
#----------Fit the semiparametric harmonic regression----------#
## Estimate the frequency (Rank-based)
phiRange <- seq(0,0.99,by=0.01) 
system.time(paraRankLS1 <- GetFitRankLS(y1,t=t,lambdaRange,phiRange)) 
system.time(paraRankLS2 <- GetFitRankLS(y2,t=t,lambdaRange,phiRange)) 
paraRankLS1[2,1]
paraRankLS2[2,1]

## Modeling given the estimated frequency
fitSHar1 <- semihregScam(y1, t, lambda0=paraRankLS1[2,1],
                          family = gaussian(), bs = "mpi") #the Gaussian case
fitSHar2 <- semihregScam(y2, t, lambda0=paraRankLS2[2,1],
                          family = gaussian(), bs = "mpi") #the Poisson case

## Fitted values at continuous time
newt <- seq(t[1],t[length(t)],0.1)
predSHar1 <- predict(fitSHar1,newt)[,2]
predSHar2 <- predict(fitSHar2,newt)[,2]

## P-values of the likelihood ratio test
pvalueRank1 <- mean(distRankLS>=paraRankLS1[1,1])
pvalueRank2 <- mean(distRankLS>=paraRankLS2[1,1])
```

The estimated frequencies using the rank-based method are `r paraRankLS1[2,1]` and `r paraRankLS2[2,1]`
for the gene YCL055W and the gene YMR198W, respectively.
The corresponding p-values are `r pvalueRank1` and `r pvalueRank2`,respectively.

## Visualize the Fitted Model
The observed time series are black dots. The fitted harmonic regression is blue plus and the
semiparametric harmonic regression is red triangle.

```{r geneFit, fig.show ='hold',fig.cap = "The observed and fitted gene expressions", fig.width=7,fig.height=3}
par(mfrow=c(1,2))

#YCL055W
plot(t,y1,type="p",pch=19,
     xlab = "time", ylab = "gene expression",
     main="YCL055W")
points(t,fitHar1$fitted.values,col="blue",pch=3)
lines(newt,predHar1,col="blue")
points(t,fitSHar1$esty,col="red",pch=6)
lines(newt,predSHar1,col="red",lty=3)

#YMR198W
plot(t,y1,type="p",pch=19,
     xlab = "time", ylab = "gene expression",
     main="YMR198W")
points(t,fitHar2$fitted.values,col="blue",pch=3)
lines(newt,predHar2,col="blue")
points(t,fitSHar2$esty,col="red",pch=6)
lines(newt,predSHar2,col="red",lty=3)

par(mfrow=c(1,1))
```

